---
title: "Homework 3"
author: "Ginevra Carbone"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

setwd("/home/ginevracoal/MEGA/Università/DSSC/semester_4/bayesian-statistics/homeworks/homework3/")

library(rstan)
library(R2WinBUGS)
library(rstanarm)
library(lubridate)
library(ggplot2)
library(bayesplot)
library(dplyr)

rstan_options(auto_write=TRUE)
```


## Exercise 1

**TH-1:** Probit model - Start from the logistic model on slide 20 `1992polls.stan` and write the probit specification of the model using the  `target+=` sintax in the model block and the `bernoulli_lpmf` function (for any doubt, see the Stan manual documentation in the Stan website). Simulate 1000 draws using the `1992polls` data and compare the results with the ones obtained using the `rstanarm::stan_glm` function.

### Probit model
```{r probit bernoulli, cache=TRUE}

# loading clean data
source("4.7_Fitting a series of regressions.R",
       echo=FALSE)

yr <- 1992
ok <- nes.year==yr & presvote<3
vote <- presvote[ok] - 1
income <- data$income[ok]

# fitting glm model
fit.glm <- rstanarm::stan_glm (vote ~ income, family=binomial(link="probit"))

# fitting stan model
data_1992 <- list(N = 1179,
                  vote = vote[is.na(vote)==FALSE],
                  income = income[is.na(income)==FALSE])
fit.rstan <- stan('probit_1992polls.stan',
              data = data_1992, 
              pars = c("alpha", "beta"),
              iter = 1000,
              chains =4)

# printing results
print(fit.glm)
print(fit.rstan)

```

Predicted values agree with the fact that the probit model is close to the logit one, with residual SD set to 1.6 instead of 1: $ - 0.88 \approx -1.40 / 1.6 $ and $0.2 \approx 0.33/1.6$.

### Comparison with different priors

```{r compare results}
# cauchy priors
fit2.rstan <- stan('cauchy_prior-1992polls.stan',
              data = data_1992, 
              pars = c("alpha", "beta"),
              iter = 1000,
              chains =4)
 
# normal non informative priors
fit3.rstan <- stan('non_inf_prior-1992polls.stan',
              data = data_1992, 
              pars = c("alpha", "beta"),
              iter = 1000,
              chains =4)

# compare results
print(fit2.rstan)
print(fit3.rstan)
```

Using a non-informative normal prior or a weakly informative cauchy prior on the parameters does not influence `Rhat`, `alpha` and `beta` predictions. We can only notice small variations in the effective sample size.

## Exercise 2

**LAB-1:** Using the most appropriate graphical tools of the `bayesplot` package, analyze the models fitted during the 4th and 5th lab sessions on the cockroaches dataset and comment.

The model we refer to is the non centered model where $\mu_b \sim \mathcal{N}(\alpha+building \,\zeta,\sigma_\mu)$ has been reparametrized with

$\mu_{b, raw} \sim\mathcal{N}(0,1)$
$\mu_b = \alpha+building \,\zeta+\sigma_\mu\,\mu_{b,raw}$

in order to solve the problem of divergent transitions.

### Prepare data
```{r prepare data, cache=TRUE, echo=FALSE}
pest_data <- readRDS('../../labs/lab4/pest_data.RDS')

N_buildings <- length(unique(pest_data$building_id))

## arrange data into a list
stan_dat_simple <- list(
  N = nrow(pest_data), 
  complaints = pest_data$complaints,
  traps = pest_data$traps,
  log_sq_foot = log(pest_data$total_sq_foot/1e4),
  live_in_super = pest_data$live_in_super
)

comp_model_NB <- stan_model('../../labs/lab4/multiple_NB_regression.stan')

fitted_model_NB <- sampling(comp_model_NB, data = stan_dat_simple)
samps_NB <- rstan::extract(fitted_model_NB)

y_rep <- samps_NB$y_rep

N_months <- length(unique(pest_data$date))

# Add some IDs for building and month
pest_data <- pest_data %>%
  mutate(
    building_fac = factor(building_id, levels = unique(building_id)),
    building_idx = as.integer(building_fac),
    ids = rep(1:N_months, N_buildings),
    mo_idx = lubridate::month(date)
  )

# Center and rescale the building specific data
building_data <- pest_data %>%
    dplyr::select(
      building_idx,
      live_in_super,
      age_of_building,
      total_sq_foot,
      average_tenant_age,
      monthly_average_rent
    ) %>%
    unique() %>%
    arrange(building_idx) %>%
    dplyr::select(-building_idx) %>%
    scale(scale=FALSE) %>%
    as.data.frame() %>%
    mutate( # scale by constants
      age_of_building = age_of_building / 10,
      total_sq_foot = total_sq_foot / 10000,
      average_tenant_age = average_tenant_age / 10,
      monthly_average_rent = monthly_average_rent / 1000
    ) %>%
    as.matrix()

stan_dat_hier <-
  with(pest_data,
        list(complaints = complaints,
             traps = traps,
             N = length(traps),
             J = N_buildings,
             log_sq_foot = log(pest_data$total_sq_foot/1e4),
             building_data = building_data[,-3],
             mo_idx = as.integer(as.factor(date)),
             K = 4,
             building_idx = building_idx
             )
       )
```

### Stan model load and fit
```{r model load fit, cache=TRUE}
# load hierarchical data
#stan_dat_hier <- #readRDS('../../labs/lab5/pest_data_longer_stan_dat.RDS')

# load and fit stan model
comp_model_NB_hier_ncp <- stan_model('../../labs/lab5/hier_NB_regression_ncp.stan')
fitted_model_NB_hier_ncp <- sampling(comp_model_NB_hier_ncp, data = stan_dat_hier, chains = 4, cores = 4)

# this is the non-centered model with no divergences
print(fitted_model_NB_hier_ncp, pars = c('sigma_mu','beta','alpha','phi','mu'))

scatter_no_divs <- mcmc_scatter(
  as.array(fitted_model_NB_hier_ncp),
  pars = c("mu[4]", 'sigma_mu'),
  transform = list('sigma_mu' = "log"),
  np = nuts_params(fitted_model_NB_hier_ncp)
)
scatter_no_divs

parcoord_no_divs <- mcmc_parcoord(
  as.array(fitted_model_NB_hier_ncp, pars = c("sigma_mu", "mu")),
  np = nuts_params(fitted_model_NB_hier_ncp)
)
parcoord_no_divs
```

### Posterior predictive check
```{r ppc density plot, cache=TRUE}
samps_NB_hier_ncp <- rstan::extract(fitted_model_NB_hier_ncp, pars = c('y_rep','inv_phi'))
y_rep <- as.matrix(fitted_model_NB_hier_ncp, pars = "y_rep")
ppc_dens_overlay(stan_dat_hier$complaints, y_rep[1:200,])
```

Here the light blue lines (there are 200 of them) represents predictive distributions, while the dark blue one is the distribution of the observed data.
We can notice that the simulated distributions resemble the observed one. 

```{r statistics distributions, cache=TRUE}
ppc_stat_grouped(
  y = stan_dat_hier$complaints,
  yrep = y_rep,
  group = pest_data$building_id,
  stat = 'mean',
  binwidth = 0.5
)

ppc_stat_grouped(
  y = stan_dat_hier$complaints,
  yrep = y_rep,
  group = pest_data$building_id,
  stat = 'sd',
  binwidth = 0.5
)
```

The distributions for both statistics show that the predictions are plausible almost over all the different buildings.


```{r ppc intervals, cache=TRUE}
ppc_intervals(
  y = stan_dat_hier$complaints,
  yrep = y_rep,
  x = stan_dat_hier$traps
) +
  labs(x = "Number of traps", y = "Number of complaints")
```

The posterior predictive intervals agree with the observed data.
Only a single observed number of complaints, the one corresponding to 7 traps, is considered unlikely by the model.

```{r std residuals, cache=TRUE}
mean_y_rep <- colMeans(y_rep)
mean_inv_phi <- mean(as.matrix(fitted_model_NB_hier_ncp, pars = "inv_phi"))
std_resid <- (stan_dat_hier$complaints - mean_y_rep) / sqrt(mean_y_rep + mean_y_rep^2*mean_inv_phi)
qplot(mean_y_rep, std_resid) + hline_at(2) + hline_at(-2)

```

Positive standard residuals seem to be as many as the negative ones, so the model does not overestimate nor underestimate the real outcomes.
The residual are not evenly distributed and present a clear pattern, so the model could be improved even more.
