---
title: "Poisson model"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
set.seed(123)

# simulate three random samples from Poisson distribution
theta <- 0.5
y20 <- rpois(20, theta)
y100 <- rpois(100, theta)
y1000 <- rpois(1000, theta)

a <- c(1, 1, 1, 2)
b <- c(.5, 2, 10, 2)
```

**1.** Plot the four prior distributions on the same graph, then add the Jeffreys’ prior.

```{r}
gamma_prior <- function(x, k){return(dgamma(x, a[k], b[k]))}
jeffreys_prior <- function(x, theta){return(1/sqrt(x))}

plot(c(), c(), xlim=c(0,3), ylim=c(0,3), xlab="", ylab="", main="Prior distributions")
for(k in 1:4) curve(gamma_prior(x,k), add=T, col=k)
curve(jeffreys_prior(x, theta), add=T, col=5)
legend(2, 2, 
       legend=c("Gamma(1,0.5)", "Gamma(1,2)", 
                "Gamma(1,10)", "Gamma(2,2)",
                "Jeffreys"),
       col=c(1:5), lty=1, cex=0.8)
```

**2.** For each simulated dataset, plot on the same graph the five posterior distributions.

```{r}
gamma_posterior <- function(x,y,k){
    return(dgamma(x,sum(y)+a[k],length(y)+b[k]))
}
jeffreys_posterior <- function(x,y,theta){
    return(dgamma(x,sum(y)+theta,length(y)))
}

plot_posteriors <- function(y){
    plot(c(), c(), xlim=c(0,2), ylim=c(0,8), main=paste("n=",length(y)), xlab="", ylab="")
    for(k in 1:5) curve(gamma_posterior(x,y,k), add=T, col=k)
    curve(jeffreys_posterior(x,y,theta), add=T, col=5)
    legend(1, 2.5, 
           legend=c("Gamma(1,0.5)", "Gamma(1,2)", "Gamma(1,10)", "Gamma(2,2)","Jeffreys"),
           col=c(1:5), lty=1, cex=0.8)
}

plot_posteriors(y20)
plot_posteriors(y100)
plot_posteriors(y1000)
```

**3.** Find the posterior expectations and the maxima a posteriori (MAP) of $\theta$.

```{r}
posterior_expectations <- function(y){
    for(k in 1:4) print((sum(y)+a[k])/(length(y)+b[k]))
    print((sum(y)+theta)/length(y))
}

posterior_expectations(y20)
posterior_expectations(y100)
posterior_expectations(y1000)

MAP <- function(y){
    for(k in 1:4) print((sum(y)+a[k]-1)/(length(y)+b[k]))
    print((sum(y)+theta)/length(y))
}

MAP(y20)
MAP(y100)
MAP(y1000)
```

**4.** Comment on plots and estimations.

**5.** Construct the 95% credible intervals, equi-tails and highest posterior density (HPD), for the three datasets using the Jeffreys’ prior.

```{r}
# fix confidence level
alpha <- 0.05

credible_intervals <- function(alpha,y) {
    return(qgamma(c(alpha*0.5,1-alpha*0.5),sum(y)+theta,length(y)))
}

credible_intervals(alpha,y20)
credible_intervals(alpha,y100)
credible_intervals(alpha,y1000)

hpd <- function(y,p){
    set.seed(123)
    y <- rgamma(1000,sum(y)+theta,length(y))
    dy<-density(y)
    md<-dy$x[dy$y==max(dy$y)]
    py<-dy$y/sum(dy$y)
    pys<-sort(py, decreasing=T)
    ths<-min(pys[cumsum(pys)< p])
    list(hpdr=range(dy$x[py>=ths]),mode=md)
}

hpd(y20,0.95)
hpd(y100,0.95)
hpd(y1000,0.95)
```
